Gravitational-wave sensitivity curves

### 2014-07-02 Sources.tex

\section{Astrophysical sources}\label{sec:sources}
All the sources described here are represented by shaded boxes in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega}. The boxes are drawn in such a way that there is a reasonable event rate for sources at a detectable SNR. Descriptions of each source are given in the following sections. Sources with short durations and sources which evolve in time over much longer timescales than our observations are drawn with flat topped boxes. Inspiralling binaries, or stochastic backgrounds of binaries, are drawn with a sloping top proportional to $f^{-2/3}$, which is result derived in section \ref{sec:stoch}. The width of the box gives the range of frequencies sources of a given type can have a while remaining at a detectable amplitude.

@@

\subsubsection{Neutron star binaries}
The inspiral and merger of a pair of neutron stars is the primary target for ground based detectors. The expected event rate for this type of source is uncertain, but estimates centre around $\rho_{\textrm{NS-NS}}=1.3\times 10^{-4}\;\textrm{Mpc}^{-3}\textrm{yr}^{-1}$, \cite{CBC}. Plotted in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega} is a box with an amplitude such that it produces a ratio $h_{c}/h_{n}=16$ for advanced LIGO at peak sensitivity and a width between $(3-300)\,\textrm{Hz}$.

\subsubsection{Supernovae}
Simulations of core collapse supernova events show that GWs between $(10^{2}-10^{3})\,\textrm{Hz}$ can be produced. The GW signal undergoes ${\cal{O}}(1)$ oscillation and hence is burst-like. \cite{2002A&A...393..523D} calculate the average maximum amplitude of GWs for a supernova at distance $r$ as
\begin{equation} h=8.9\times 10^{-21}\left( \frac{10 \,\textrm{kpc}}{r} \right) \; .\end{equation}
The event rate for supernova is approximately $\rho_{\textrm{SN}}=5\times10^{-4}\,\textrm{Mpc}^{-3}\textrm{yr}^{-1}$. The box plotted in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega} corresponds to a distance $r=3 \,\textrm{Mpc}$ with the frequency range quoted above. The LIGO and VIRGO detectors have already placed bounds on the event rate for these sources, \cite{Bursts}.

@@

\subsubsection{Massive black hole binaries}
Space based detectors will be sensitive to equal mass mergers in the range $(10^{4}-10^{7})\,\Msun$. Predictions of the event rate for these mergers range from ${\cal{O}}(3-300)\,\textrm{yr}^{-1}$ for LISA with SNRs of up to 10000. The uncertainty in the rate reflects our uncertainty in the growth mechanisms of the supermassive black hole population. Plotted in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega} is a box with a ratio $h_{c}/h_{n}=100$ for eLISA at its peak sensitivity. The range of frequencies plotted is $(3\times 10^{-4}-3\times 10^{-1})\,\textrm{Hz}$, this corresponds to circular binaries in the mass range quoted above.

@@

Galactic binaries divide into two classes; the unresolvable and the resolvable galactic binaries. The unresolvable binaries overlap to form a stochastic background as discussed in section \ref{sec:stoch}. Plotted in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega} is the estimate of this background 
due to \cite{Nelemans} where an observation time of one year has been assumed
\begin{equation} h_{c}(f)= 5\times 10^{-21} \left(\frac{f}{10^{-3}\,\textrm{Hz}}\right)^{-2/3} \; . \end{equation}
Estimates for the event rate of resolvable binaries centre around ${\cal{O}}(10^{3})$ events for eLISA. The box plotted in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega} has an ratio $h_{c}/h_{n}=50$ for eLISA at its peak sensitivity, the frequency range of the box is $\left(3\times10^{-4}\right.$--$\left.10^{-2}\right)~\mathrm{Hz}$, based on visual inspection of Monte Carlo population simulation results presented in \cite{Amaro-Seoane-et-al}.

\subsubsection{Extreme mass ratio inspirals}
EMRI events occur when a compact stellar mass object inspirals into a supermassive black hole. There is extreme uncertainty in the event rate for EMRIs due to the poorly constrained astrophysics in galactic centres; the best guess estimate is ${\mathcal{O}}(10)$ events per year with eLISA with SNR $\ge 20$. The box plotted in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega} has a characteristic strain of $h_{c}=3\times 10^{-20}$ at $10^{-2}~\mathrm{Hz}$ which corresponds to a $10\Msun$ BH inspiralling into a $10^{6}\Msun$ black hole at a luminosity distance of $1~\mathrm{Gpc}$. The frequency width of the box is somewhat unknown, EMRI events can occur into a black hole of any mass, and hence EMRIs can occur at any frequency. However there is substantial uncertainty in the black hole mass function, so the box in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega} is drawn with a width comparable to that of the LISA sensitivity curve.

@@

Supermassive black hole binaries at higher frequencies are inspiralling faster and hence there are fewer of them per frequency bin. At a certain frequency, these sources will cease to be a background and become individually resolvable. It is currently unclear whether PTAs will detect an individual binary or a stochastic background first. Plotted in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega} for the unresolvable background is a third of the current limit with a cut off frequency of $f=1~\mathrm{yr}^{-1}$ which is suggested by Monte Carlo population studies, \cite{SesanaVecchioColancino}. For the resolvable sources the amplitude of the current limit is plotted between $\left(3\times 10^{-9}\right.$--$\left.3\times 10^{-7}\right)~\mathrm{Hz}$.

@@

In addition to the sources above, early Universe processes, such as inflation \citep{Grishchuk2005} or a first-order phase transition \citep{Binetruy2012}, could have created GWs. More speculatively, it has been hypothesised that cosmic strings could also be a potential source \citep{Damour2005,Binetruy2012}. These relic GWs allow us to explore energy scales far beyond those accessible by other means, providing insight into new and exotic physics. The excitement surrounding the tentative discovery by BICEP2 of the imprint of primordial GWs (generated during inflation) in the cosmic microwave background \citep{Ade2014}, and the subsequent flurry of activity, has shown the scientific potential of such cosmological GWs. These GW signals are so alluring because they probe unknown physics; this also makes them difficult to predict. Cosmological stochastic backgrounds have been predicted across a range of frequencies with considerable variation in amplitude. As a consequence of this uncertainty, although we could learn much from measuring these signals, we have not included them amongst the sources shown in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega}.

>>>>

\section{Astrophysical sources}\label{sec:sources}
All the sources described here are represented by shaded boxes in \ref{app:a}. The boxes are drawn in such a way that there is a reasonable event rate for sources at a detectable SNR. Descriptions of each source are given in the following sections. Sources with short durations and sources which evolve in time over much longer timescales than our observations are drawn with flat topped boxes. Inspiralling binaries, or stochastic backgrounds of binaries, are drawn with a sloping top proportional to $f^{-2/3}$, which is result derived in section \ref{sec:stoch}. The width of the box gives the range of frequencies sources of a given type can have a while remaining at a detectable amplitude.

@@

\subsubsection{Neutron star binaries}
The inspiral and merger of a pair of neutron stars is the primary target for ground based detectors. The expected event rate for this type of source is uncertain, but estimates centre around $\rho_{\textrm{NS-NS}}=1.3\times 10^{-4}\;\textrm{Mpc}^{-3}\textrm{yr}^{-1}$, \cite{CBC}. Plotted in \ref{app:a} is a box with an amplitude such that it produces a ratio $h_{c}/h_{n}=16$ for advanced LIGO at peak sensitivity and a width between $(3-300)\,\textrm{Hz}$.

\subsubsection{Supernovae}
Simulations of core collapse supernova events show that GWs between $(10^{2}-10^{3})\,\textrm{Hz}$ can be produced. The GW signal undergoes ${\cal{O}}(1)$ oscillation and hence is burst-like. \cite{2002A&A...393..523D} calculate the average maximum amplitude of GWs for a supernova at distance $r$ as
\begin{equation} h=8.9\times 10^{-21}\left( \frac{10 \,\textrm{kpc}}{r} \right) \; .\end{equation}
The event rate for supernova is approximately $\rho_{\textrm{SN}}=5\times10^{-4}\,\textrm{Mpc}^{-3}\textrm{yr}^{-1}$. The boxes plotted in \ref{app:a} correspond to a distance $r=3 \,\textrm{Mpc}$ with the frequency range quoted above. The LIGO and VIRGO detectors have already placed bounds on the event rate for these sources, \cite{Bursts}.

@@

\subsubsection{Massive black hole binaries}
Space based detectors will be sensitive to equal mass mergers in the range $(10^{4}-10^{7})\,\Msun$. Predictions of the event rate for these mergers range from ${\cal{O}}(3-300)\,\textrm{yr}^{-1}$ for LISA with SNRs of up to 10000. The uncertainty in the rate reflects our uncertainty in the growth mechanisms of the supermassive black hole population. Plotted in \ref{app:a} are boxes with a ratio $h_{c}/h_{n}=100$ for eLISA at its peak sensitivity. The range of frequencies plotted is $(3\times 10^{-4}-3\times 10^{-1})\,\textrm{Hz}$, this corresponds to circular binaries in the mass range quoted above.

@@

Galactic binaries divide into two classes: the unresolvable and the resolvable galactic binaries. The unresolvable binaries overlap to form a stochastic background as discussed in section \ref{sec:stoch}. Plotted in \ref{app:a} is the estimate of this background due to \cite{Nelemans} where an observation time of one year has been assumed
\begin{equation} h_{c}(f)= 5\times 10^{-21} \left(\frac{f}{10^{-3}\,\textrm{Hz}}\right)^{-2/3} \; . \end{equation}
Estimates for the event rate of resolvable binaries centre around ${\cal{O}}(10^{3})$ events for eLISA. The boxes plotted in \ref{app:a} have a ratio $h_{c}/h_{n}=50$ for eLISA at its peak sensitivity. The frequency range of the box is $\left(3\times10^{-4}\right.$--$\left.10^{-2}\right)~\mathrm{Hz}$, based on visual inspection of Monte Carlo population simulation results presented in \cite{Amaro-Seoane-et-al}.

\subsubsection{Extreme mass ratio inspirals}
EMRI events occur when a compact stellar mass object inspirals into a supermassive black hole. There is extreme uncertainty in the event rate for EMRIs due to the poorly constrained astrophysics in galactic centres; the best guess estimate is ${\mathcal{O}}(10)$ events per year with eLISA with SNR $\ge 20$. The boxes plotted in \ref{app:a} have a characteristic strain of $h_{c}=3\times 10^{-20}$ at $10^{-2}~\mathrm{Hz}$, which corresponds to a $10\Msun$ BH inspiralling into a $10^{6}\Msun$ black hole at a luminosity distance of $1~\mathrm{Gpc}$. The frequency width of the box is somewhat unknown; EMRI events can occur into a black hole of any mass, and hence EMRIs can occur at any frequency. However there is substantial uncertainty in the black hole mass function, so the boxes in \ref{app:a} are drawn with a width comparable to that of the LISA sensitivity curve.

@@

Supermassive black hole binaries at higher frequencies are inspiralling faster and hence there are fewer of them per frequency bin. At a certain frequency, these sources will cease to be a background and become individually resolvable. It is currently unclear whether PTAs will detect an individual binary or a stochastic background first. Plotted in \ref{app:a} for the unresolvable background is a third of the current limit with a cut off frequency of $f=1~\mathrm{yr}^{-1}$ which is suggested by Monte Carlo population studies, \cite{SesanaVecchioColancino}. For the resolvable sources the amplitude of the current limit is plotted between $\left(3\times 10^{-9}\right.$--$\left.3\times 10^{-7}\right)~\mathrm{Hz}$.

@@

In addition to the sources above, early Universe processes, such as inflation \citep{Grishchuk2005} or a first-order phase transition \citep{Binetruy2012}, could have created GWs. More speculatively, it has been hypothesised that cosmic strings could also be a potential source \citep{Damour2005,Binetruy2012}. These relic GWs allow us to explore energy scales far beyond those accessible by other means, providing insight into new and exotic physics. The excitement surrounding the tentative discovery by BICEP2 of the imprint of primordial GWs (generated during inflation) in the cosmic microwave background \citep{Ade2014}, and the subsequent flurry of activity, has shown the scientific potential of such cosmological GWs. These GW signals are so alluring because they probe unknown physics; this also makes them difficult to predict. Cosmological stochastic backgrounds have been predicted across a range of frequencies with considerable variation in amplitude. As a consequence of this uncertainty, although we could learn much from measuring these signals, we have not included them amongst the sources shown in \ref{app:a}.
